Impact of Roe v. Wade’s Overturn: Abortion Restrictions and Infant Mortality in the U.S. Using a Difference-in-Differences Analysis

### 2024-12-01 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper with check

---
title: "Impact of Roe v. Wade’s Overturn: Abortion Restrictions and Infant Mortality in the U.S. Using a Difference-in-Differences Analysis"
subtitle: "Infant death rates rose by 0.285 per 1,000 live births in abortion-restricted states compared to protected states."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion](https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "This study investigates the impact of the Dobbs decision and abortion bans on U.S. infant mortality rates using 2021–2022 data. A Difference-in-Differences (DID) analysis shows a 0.285 increase in infant deaths per 1,000 live births in states enforcing bans, with amplified effects post-injunction. Maternal age and race are key factors, with Black mothers experiencing higher mortality rates. These findings highlight the public health impact of restrictive abortion policies and the need for targeted interventions to address disparities. This study uses the latest available data and rigorous modeling to explore the link between reproductive rights and infant health."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Instead of examining the impact of abortion laws over a broader time
periods like prior research, our study focuses on the effect of abortion
bans on infant mortality rates in the United States specifically from
2021 to 2022, with an emphasis on changes following the 2022 Dobbs
decision. Using a Difference-in-Differences (DID) approach, we assess
the interaction between abortion legality in states and the timing of
the injunction implemented to understand its influence on infant health
outcomes. The study utilizes observational data from the Centers for
Disease Control and Prevention’s WONDER database, captures infant
mortality trends across states with varying levels of abortion
restrictiveness. Key predictors include the mother’s state of residence,
race, age, and the year of infant death, enabling a detailed examination
of the effects of restrictive abortion laws on public health.

@@

We conduct our infant death rate observational data analysis using the R
programming language [@citeR]. Our dataset, obtained from the CDC
Wonder[@cdc_wonder_2024], based on data as of December 2022, provides a
detailed overview of the Infant Death Rate across the USA. CDC WONDER
(Wide-ranging ONline Data for Epidemiologic Research) is an online
system providing public health professionals and the public with access
to a wide range of the Centers for Disease Control and Prevention (CDC)
epidemiologic data and resources. Following the guidelines outlined in
@tellingstories, we analyze the impact of abortion injunctions on infant
death rates using the Difference-in-Difference method. This approach
utilizes the interaction between the legality of abortion in the
mother’s state of residence and the timing of the infant’s death as the
predictor variable. Additionally, we include control variables such as
infant year of death, the mother’s state of residence, race, and age to
account for potential confounding factors.

@@

The `death rate` refers to the number of deaths of infants under one
year of age per 1,000 live births in a specified population during a
given time period. It serves as a key indicator of a population’s
overall health and access to healthcare. Rates labeled as "Unreliable"
occur when there are fewer than 20 deaths in the numerator, as these
figures do not meet the National Center for Health Statistics (NCHS)
standards for reliability or precision. After cleaning the dataset, the "Unreliable" label is converted into a separate variable named
`reliable`.

@@

Abortion legality (named `abortion_illegal` in the dataset) is a newly
created variable in our dataset, which refers to the legal status of
abortion within a given state. It is represented as a binary variable,
where 1 indicates that abortion is illegal in the state, and 0 indicates
that abortion is legal. This classification reflects the policies and
laws governing abortion access in each state.

@@

```{r}
#| label: fig-map
#| fig-cap: >
#|   Abortion Legality by State: After June 2022, abortion is prohibited in the
#|   following states: Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana,
#|   Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee,
#|   Texas, West Virginia, and Wisconsin.
#| echo: false
ggplot() +
  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group),
    color = "black") +
  scale_fill_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Legal", "Illegal"),
    name = "Abortion Legal Or Not"
  )+
  theme_void() +
  labs(
    caption = "Source: Center for Reproductive Rights\nDate as of Nov.19 2024"
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-rateandstate
#| fig-cap: >
#|   Infant Death Rates Over Time: Infant death rates (per 1,000 live births)
#|   show a slight increase in states where abortion became illegal following
#|   June 2022, whereas rates in states with legal abortion remained relatively
#|   stable. 
#| echo: false
#| message: false
analysis_data|>
  mutate(illegal = if_else( abortion_illegal== 1, "Illegal State", "Legal State"))|>
  ggplot(aes(x = date, y = death_rate,colour = illegal)) +
  geom_point(alpha = 0.5)+
  geom_smooth() +
  labs(x = "Date of Infant Death", y = "Infant Death Rate \n(per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S. Supreme Court overturned Roe v. Wade.") +
  facet_wrap(vars(illegal), nrow = 2)+
  geom_vline(xintercept = as.Date("2022-06-01"), linetype = "dashed")  +
  scale_x_date(
    limits = c(as.Date("2021-01-01"), as.Date("2022-12-01")), 
    date_breaks = "3 months",  
    date_labels = "%Y-%m" 
  )+
  scale_color_brewer(palette = "Set1")+
  theme_classic() +
  theme(legend.position = "bottom")
```

@@

```{r}
#| label: tbl-interact
#| tbl-cap: >
#|   Average Infant Death Rate by Abortion Legality and Injunction Timing: 
#|   States categorized as "Illegal Abortion and After Injunction" ('1') 
#|   exhibit a higher average infant death rate (per 1,000 live births) compared 
#|   to states in other categories, including "Legal Abortion and Before/After 
#|   Injunction" or "Illegal Abortion and Before Injunction" ('0'). 
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
  group_by(illegal_after_injunction)|>rename(`Abortion Restriction Post-Overturn`=illegal_after_injunction)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1))

@@

Another created variable `after_injunction` indicates whether the date
of infants' death falls after the implementation of the abortion
injunctions following the overturn of Roe v. Wade in June 2022. It is
coded as a binary variable, where 1 represents dates after June 2022,
signifying the post-injunction period, and 0 represents dates before
June 2022, indicating the pre-injunction period.

@@

@tbl-summary shows that 50.7% of infant deaths occurred in 2022, slightly higher than in 2021 (49.3%). The majority of cases were among
mothers aged 25–29 (37.8%), followed by those aged 30–34 (32.5%), and
20–24 (27.2%), with the lowest occurrence in mothers aged 15–19 (2.6%).
By race, White mothers accounted for the highest percentage (75.7%),
followed by Black or African American mothers (24.0%), and Asian mothers
(0.3%).

@@

The Year of Death variable represents the calendar year in which the
infant death occurred, specifically focusing on the years 2021 and 2022
in this study. This variable is essential for analyzing temporal trends
and identifying patterns or shifts in infant mortality rates over this
essential period. By controlling the year of death, we can closely
examine the impact of significant policy changes, such as the
overturning of Roe v. Wade in 2022, as well as other temporal factors on
infant health outcomes.

@@

The self-identified race of the mother was recorded as a single race
category. In this dataset, the available categories are: Asian, Black or
African American, and White. This classification facilitates the
analysis of maternal and infant health outcomes across these racial
groups.

@@

@@

The state of the mother’s legal residence at the time of birth. This
data is categorized by the mother’s state of residence and excludes
territories.

```{r}
#| label: fig-choromap
#| fig-cap: >
#|   Infant Death Rates Across U.S. States (per 1,000 Live Births): 
#|   The infant mortality rate is significantly higher in the Southeastern states, 
#|   with Alabama and Mississippi reporting the highest rates. In contrast, 
#|   states like California and New York exhibit notably lower rates.
#| echo: false
death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
death_full <- us_map|>
   full_join(death, c("region"="state"))

@@

Define $y_i$ as as the infant death rate per 1,000 live births, and
$\gamma_j$ as the random state effect. $$
\begin{aligned}
y_i|\mu_i,\sigma &\sim \mbox{Normal}(\mu_i, \sigma)\\
\text{First Model}: \mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction} + \beta_2\cdot\text{Abortion Illegal}\\&+\beta_3\cdot\text{After Injunction}\cdot\text{Abortion Illegal}_i+\gamma_j\\
\text{Second Model}:\mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction}_i + \beta_2\cdot\text{Abortion Illegal}_i\\&+ \beta_3\cdot\text{After Injunction}_i\cdot\text{Abortion Illegal}_i\\&+\beta_4\cdot\text{Year of Death}_i+\beta_5\cdot\text{Age of Mother}_i+\beta_6\cdot\text{Mother's Single Race}_i+\gamma_j\\
\beta_0 &\sim \mbox{Normal}(0, 2.5)\\
\beta_1 &\sim \mbox{Normal}(0, 2.5)\\
\beta_2 &\sim \mbox{Normal}(0, 2.5)\\
\beta_3 &\sim \mbox{Normal}(0, 2.5)\\
\beta_4 &\sim \mbox{Normal}(0, 2.5)\\
\beta_5 &\sim \mbox{Normal}(0, 2.5)\\
\beta_6 &\sim \mbox{Normal}(0, 2.5)\\
\gamma_j&\sim \text{Normal}(0, \sigma_j^2)\\
\sigma &\sim \mbox{Exponential}(1)
\end{aligned}$$

The models are run in R (R Core Team 2023) using the `rstanarm` package
of @rstanarm. The default priors from `rstanarm` are used for both GLM
Bayesian models. We run the model in R [@citeR] using the `rstanarm`
package of @rstanarm. We use the default priors from `rstanarm` used for
both Gaussian Bayesian models.

@@

We use a Bayesian approach to enhance the Difference-in-Differences
(DID) method, which allows us to incorporate prior knowledge and better
account for uncertainty. With the `rstanarm` package, we apply weakly
informative priors—such as a Normal prior centered at zero with moderate
spread ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential
prior (rate = 1) for variability parameters. These priors automatically adjust to the scale of the data, improving model stability and ensuring they complement rather than dominate the observed data. This approach
helps us center estimates realistically, account for hidden differences
between groups, and produce reliable results while assuming trends
between treated and control groups are similar before treatment.

@@

The results from the two Bayesian models are summarized in
@tbl-modelresults. The interaction term between after_injunction and
abortion_illegal is positive and statistically significant (0.111 in
first model and 0.280 in the second model), indicating a change in infant
mortality rates associated with the combination of abortion bans and the
post-injunction period.

@@

The models performed well, with low RMSE values and high $R^2$ values,
indicating that the predictors account for a substantial portion of the
variance in infant mortality rates.

@@

```{r}
#| label: fig-eventstudy
#| fig-cap: >
#|   Event Study Analysis of the Impact of Abortion Bans on Infant Mortality Rates in the United   States: 
#|   Evidence Using June 2022 as the Treatment Date and Pre-Treatment (-1, May 2022) as the Reference Period.
#| warning: false
#| echo: false
#| eval: true

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheck
#| layout-ncol: 1
#| fig-cap: "Posterior Predictive Check: Examining how the model for infant death rate fits, and is affected by, the data"
#| fig-subcap: ["First Model Posterior prediction check", "Second Model Posterior prediction check" ]

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-posteriorvsprior1
#| fig-cap: "Comparing Prior and Posterior Distributions: Examining how the first model for infant death rate fits, and is affected by, the data"
posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-posteriorvsprior2
#| fig-cap: "Comparing Prior and Posterior Distributions: Examining how the second model for infant death rate fits, and is affected by, the data"
#| 
posterior_vs_prior(second_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

@@

@tbl-priorsummary shows the prior summaries for the two models, `first_model` and `second_model`, indicating that both use weakly
informative priors. For the intercept, a normal prior with a mean of 0
and a standard deviation of 2.5 was specified, which adjusts to a scale
of 12 after centering the predictors. The coefficients for both models
are also assigned normal priors with a mean of 0 and a scale of 2.5,
which are adjusted based on the predictors, resulting in scales such as
\[26.84, 28.43, 45.17\] for the first model and \[26.84, 28.43,
24.78,...\] for the second model. The auxiliary parameter (sigma)
follows an exponential prior with a rate of 1, adjusted to 0.2, while a
decov prior regularizes the covariance structure in both models.
Although the prior structure is similar, the second model incorporates
more predictors, leading to a larger number of coefficients with
slightly adjusted scales. These priors are designed to be weakly
informative, allowing the data to primarily guide the posterior
estimates while avoiding extreme parameter values.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-area
#| layout-ncol: 1
#| fig-cap: "95% credibility intervals for fitted models"
#| fig-subcap: ["First Model", "Second Model"]

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay1
#| fig-cap: "First Model: Checking the convergence of the MCMC algorithm for the infant death rate model"
#| fig-subcap: ["First Model Trace plot", "First Model Rhat plot"]
#| layout-ncol: 1

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay2
#| fig-cap: "Second Model: Checking the convergence of the MCMC algorithm for the infant death rate model"
#| fig-subcap: ["Second Model Trace plot", "Second Model Rhat plot"]
#| layout-ncol: 1
plot(second_model, "trace")

>>>>

---
title: "Impact of Roe v. Wade’s Overturn: Abortion Restrictions and Infant Mortality in the U.S. Using a Difference-in-Differences Analysis"
subtitle: "Infant death rates rose by 0.285 per 1,000 live births in abortion-restricted states compared to protected states."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion](https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "This study investigates the impact of the Dobbs decision and abortion bans on U.S. infant mortality rates using 2021–2022 data. A Difference-in-Differences (DID) analysis shows a 0.285 increase in infant deaths per 1,000 live births in states enforcing bans, with amplified effects post-injunction. Maternal age and race are key factors, with Black mothers experiencing higher infant mortality rates. These findings highlight the public health impact of restrictive abortion policies and the need for targeted interventions to address disparities. This study uses the latest available data and rigorous modeling to explore the link between reproductive rights and infant health."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Instead of examining the impact of abortion laws over a broader time
periods like prior research, our study focuses on the effect of abortion
bans on infant mortality rates in the United States specifically from
2021 to 2022, with an emphasis on changes following the 2022 Dobbs
decision. Using a Difference-in-Differences (DID) approach, we assess
the interaction between abortion legality in states and the timing of
the injunction implemented to understand its influence on infant health
outcomes. The study utilizes observational data from the Centers for
Disease Control and Prevention’s WONDER database [@cdc_wonder_2024], captures infant
mortality trends across states with varying levels of abortion
restrictiveness. Key predictors include the mother’s state of residence,
race, age, and the year of infant death, enabling a detailed examination
of the effects of restrictive abortion laws on public health.

@@

We conduct our infant death rate observational data analysis using the R
programming language [@citeR]. Our dataset, obtained from the CDC
Wonder[@cdc_wonder_2024], based on data as of December 2022, provides detailed state-level information on infant death rates in the United States, serving as the foundation for our study of the impact of abortion injunctions. CDC WONDER
(Wide-ranging ONline Data for Epidemiologic Research) is an online
system providing public health professionals and the public with access
to a wide range of the Centers for Disease Control and Prevention (CDC)
epidemiologic data and resources. This dataset offers detailed state-level information on infant death rates in the United States, serving as the foundation for our study of the impact of abortion injunctions.
Following the guidelines outlined in
@tellingstories, we analyze the impact of abortion injunctions on infant
death rates using the Difference-in-Difference method. This approach, shown in @sec-model, utilizes the main effects and the interaction between the legality of abortion in the
mother’s state of residence and the timing of the infant’s death as the predictor variable. Additionally, we include control variables, infant year of death, race, and age to account for potential confounding factors, and the mother’s state of residence as a random variable.

We selected the CDC WONDER dataset for its combination of granularity, consistency, and accessibility, making it suitable for a nationwide analysis. Alternative datasets lacked either the breadth or the standardization required for a study of this scale. Additionally, CDC WONDER’s public availability enhances transparency and allows for independent verification, ensuring the reliability and reproducibility of our research.

@@

The death rate (`death rate`) refers to the number of deaths of infants under one
year of age per 1,000 live births in a specified population during a
given time period. It serves as a key indicator of a population’s
overall health and access to healthcare. Rates labeled as "Unreliable"
occur when there are fewer than 20 deaths in the numerator, as these
figures do not meet the National Center for Health Statistics (NCHS)
standards for reliability or precision. After cleaning the dataset, the "Unreliable" label is converted into a separate variable named `reliable` (This variable is available in the `analysis data` for reference but was not included in the model analysis.).

@@

### Abortion Legality

Abortion legality (`abortion_illegal`) is a newly
created variable in our dataset, which refers to the legal status of
abortion within a given state. It is represented as a binary variable,
where 1 indicates that abortion is illegal in the state, and 0 indicates
that abortion is legal. This classification reflects the policies and
laws governing abortion access in each state.

@@

```{r}
#| label: fig-map
#| fig-cap: >
#|   Abortion Legality by State: After June 2022, abortion is prohibited in the
#|   following states: Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana,
#|   Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee,
#|   Texas, West Virginia, and Wisconsin.
#| echo: false
#| fig-pos: "H"
ggplot() +
  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group),
    color = "black") +
  scale_fill_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Legal", "Illegal"),
    name = "State is Illegal for Abortion or Not"
  )+
  theme_void() +
  labs(
    caption = "Source: Center for Reproductive Rights\nDate as of Nov.19 2024"
  ) +
  theme(legend.position = "bottom")
```

### After Injunction

Another created variable After Injunction (`after_injunction`) indicates whether the date
of infants' death falls after the implementation of the abortion
injunctions following the overturn of Roe v. Wade in June 2022. It is
coded as a binary variable, where 1 represents dates after June 2022,
signifying the post-injunction period, and 0 represents dates before
June 2022, indicating the pre-injunction period.

@@

```{r}
#| label: fig-rateandstate
#| fig-cap: >
#|   Infant Death Rates Over Time: Infant death rates (per 1,000 live births)
#|   show a slight increase in states where abortion became illegal following
#|   June 2022, whereas rates in states with legal abortion remained relatively
#|   stable. 
#| echo: false
#| message: false
#| fig-pos: "H"
analysis_data|>
  mutate(illegal = if_else( abortion_illegal== 1, "Illegal State", "Legal State"))|>
  ggplot(aes(x = date, y = death_rate,colour = illegal)) +
  geom_point(alpha = 0.5)+
  geom_smooth() +
  labs(x = "Date of Infant Death", y = "Infant Death Rate \n(per 1,000 live births)",color="State is Illegal for Abortion or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S. Supreme Court overturned Roe v. Wade.") +
  facet_wrap(vars(illegal), nrow = 2)+
  geom_vline(xintercept = as.Date("2022-06-01"), linetype = "dashed")  +
  scale_x_date(
    limits = c(as.Date("2021-01-01"), as.Date("2022-12-01")), 
    date_breaks = "3 months",  
    date_labels = "%Y-%m" 
  )+
  scale_color_brewer(palette = "Set1")+
  theme_classic() +
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-interact
#| tbl-cap: >
#|   Average Infant Death Rate by Abortion Legality and Injunction Timing: 
#|   States categorized as "Illegal Abortion and After Injunction" ('1') 
#|   exhibit a higher average infant death rate (per 1,000 live births) compared 
#|   to states in other categories, including "Legal Abortion and Before/After 
#|   Injunction" or "Illegal Abortion and Before Injunction" ('0'). 
#| echo: false
#| warning: false
#| message: false

analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
  group_by(illegal_after_injunction)|>rename(`Abortion Restriction Post-Overturn`=illegal_after_injunction)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1))

@@

@@

@tbl-summary shows the summary statistics of control variables. 50.7% of infant deaths occurred in 2022, slightly higher than in 2021 (49.3%). The majority of cases were among
mothers aged 25–29 (37.8%), followed by those aged 30–34 (32.5%), and
20–24 (27.2%), with the lowest occurrence in mothers aged 15–19 (2.6%).
By race, White mothers accounted for the highest percentage (75.7%),
followed by Black or African American mothers (24.0%), and Asian mothers
(0.3%).

@@

The Year of Death (`year_of_death`) represents the calendar year in which the
infant death occurred, specifically focusing on the years 2021 and 2022
in this study. This variable is essential for analyzing temporal trends
and identifying patterns or shifts in infant mortality rates over this
essential period. By controlling the year of death, we can closely
examine the impact of significant policy changes, such as the
overturning of Roe v. Wade in 2022, as well as other temporal factors on
infant health outcomes.

@@

Mother's Single Race (`mothers_single_race`) refers to the self-identified race of the mother, which was recorded as a single race category. In this dataset, the available categories are Asian, Black or African American, and White. This classification facilitates the analysis of maternal and infant health outcomes across these racial groups.

@@

@tbl-motherrace represents the average infant death rates (per 1,000 live births) by maternal race. Black or African American mothers have the highest rate at 14.5, compared to 5.8 for White mothers and 5.4 for Asian mothers. These differences indicate significant disparities in infant mortality outcomes across racial groups.

@@

State (`state`) referes to the state of the mother’s legal residence at the time of birth. This
data is categorized by the mother’s state of residence and excludes territories.

```{r}
#| label: fig-choromap
#| fig-cap: >
#|   Infant Death Rates Across U.S. States (per 1,000 Live Births): 
#|   The infant mortality rate is significantly higher in the Southeastern states, 
#|   with Alabama and Mississippi reporting the highest rates. In contrast, 
#|   states like California and New York exhibit notably lower rates.
#| echo: false
#| fig-pos: "H"
death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
death_full <- us_map|>
   full_join(death, c("region"="state"))

@@

Define $y_i$ as as the infant death rate per 1,000 live births, and
$\gamma_j$ as the random state effect for state j. 
$$
\begin{aligned}
y_i|\mu_i,\sigma &\sim \mbox{Normal}(\mu_i, \sigma)\\
\text{First Model}: \mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction} + \beta_2\cdot\text{Abortion Illegal}\\&+\beta_3\cdot\text{After Injunction}\cdot\text{Abortion Illegal}_i+\gamma_j\\
\text{Second Model}:\mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction}_i + \beta_2\cdot\text{Abortion Illegal}_i\\&+ \beta_3\cdot\text{After Injunction}_i\cdot\text{Abortion Illegal}_i\\&+\beta_4\cdot\text{Year of Death}_i+\beta_5\cdot\text{Age of Mother}_i+\beta_6\cdot\text{Mother's Single Race}_i+\gamma_j\\
\beta_0 &\sim \mbox{Normal}(0, 2.5)\\
\beta_1 &\sim \mbox{Normal}(0, 2.5)\\
\beta_2 &\sim \mbox{Normal}(0, 2.5)\\
\beta_3 &\sim \mbox{Normal}(0, 2.5)\\
\beta_4 &\sim \mbox{Normal}(0, 2.5)\\
\beta_5 &\sim \mbox{Normal}(0, 2.5)\\
\beta_6 &\sim \mbox{Normal}(0, 2.5)\\
\gamma_j&\sim \text{Normal}(0, \sigma_j^2)\\
\sigma &\sim \mbox{Exponential}(1)
\end{aligned}
$$

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm` used for both Gaussian Bayesian models.

@@

We use a Bayesian approach to enhance the Difference-in-Differences
(DID) method, which allows us to incorporate prior knowledge and better
account for uncertainty. With the `rstanarm` package, we apply weakly
informative priors—such as a Normal prior centered at zero with moderate
spread ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential
prior (rate = 1) for variability parameters. By using `autoscale=TRUE`, these priors automatically adjust to the scale of the data, improving model stability and ensuring they complement rather than dominate the observed data. This approach
helps us center estimates realistically, account for hidden differences
between groups, and produce reliable results while assuming trends
between treated and control groups are similar before treatment.

@@

The results from the two Bayesian models are summarized in
@tbl-modelresults. The interaction term between after_injunction and
abortion_illegal is positive and significant (0.111 in
first model and 0.280 in the second model), indicating a change in infant
mortality rates associated with the combination of abortion bans and the
post-injunction period.

@@

Both models exhibit low RMSE values and high $R^2$, indicating a strong overall fit and that the predictors explain much of the variance in infant mortality rates. The second model, with its additional covariates, provides a more detailed and adjusted view of the intervention’s effects, while the first model offers a simpler and more direct perspective. The choice between models depends on the balance between interpretability and precision.

@@

```{r}
#| label: fig-eventstudy
#| fig-cap: >
#|   Event Study Analysis of the Impact of Abortion Bans on Infant Mortality Rates in the United   States: 
#|   Evidence Using June 2022 as the Treatment Date and Pre-Treatment (-1, May 2022) as the Reference Period.
#| warning: false
#| echo: false
#| eval: true
#| fig-pos: "H"

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheck
#| layout-ncol: 1
#| fig-cap: "Posterior Predictive Check: Examining how the model for infant death rate fits, and is affected by, the data"
#| fig-subcap: ["First Model Posterior prediction check", "Second Model Posterior prediction check" ]
#| fig-pos: "H"

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-posteriorvsprior1
#| fig-cap: "Comparing Prior and Posterior Distributions: Examining how the first model for infant death rate fits, and is affected by, the data"
#| fig-pos: "H"
posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-posteriorvsprior2
#| fig-cap: "Comparing Prior and Posterior Distributions: Examining how the second model for infant death rate fits, and is affected by, the data"
#| fig-pos: "H"
posterior_vs_prior(second_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

@@

@tbl-priorsummary shows the prior summaries for the two models, indicating that both use weakly
informative priors. For the intercept, a normal prior with a mean of 0
and a standard deviation of 2.5 was specified, which adjusts to a scale
of 12 after centering the predictors. The coefficients for both models
are also assigned normal priors with a mean of 0 and a scale of 2.5,
which are adjusted based on the predictors, resulting in scales such as
\[26.84, 28.43, 45.17\] for the first model and \[26.84, 28.43,
24.78,...\] for the second model. The auxiliary parameter (sigma)
follows an exponential prior with a rate of 1, adjusted to 0.2, while a
decov prior regularizes the covariance structure in both models.
Although the prior structure is similar, the second model incorporates
more predictors, leading to a larger number of coefficients with
slightly adjusted scales. These priors are designed to be weakly
informative, allowing the data to primarily guide the posterior
estimates while avoiding extreme parameter values.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-area
#| layout-ncol: 2
#| fig-cap: "95% credibility intervals for fitted models"
#| fig-subcap: ["First Model", "Second Model"]
#| fig-pos: "H"

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay1
#| fig-cap: "First Model: Checking the convergence of the MCMC algorithm for the infant death rate model"
#| fig-subcap: ["First Model Trace plot", "First Model Rhat plot"]
#| layout-ncol: 1
#| fig-pos: "H"

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay2
#| fig-cap: "Second Model: Checking the convergence of the MCMC algorithm for the infant death rate model"
#| fig-subcap: ["Second Model Trace plot", "Second Model Rhat plot"]
#| layout-ncol: 1
#| fig-pos: "H"
plot(second_model, "trace")
